The natural history of scavenging in vertebrates

### 2016-03-16 natural history of scavenging.tex
Added equations templates
Minor updates (mainly dates and a couple of comments)

\documentclass[a4paper,12pt]{article}
\usepackage{ecography}
\usepackage{lmodern}

@@

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today. Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Moreover, having a panoramic view means being able to gather a wealth of information from other foragers, be they conspecifics or other species \citep{jackson2008effect}. Again, returing to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}. These behaviours allow them forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \citep{KaneVul}.

@@

Avian flight originates in the Jurassic Period, conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Triassic Period. Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres) and look to have engaged in soaring flight \citep{witton2010size}.  Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Large body size confers substantial dominance benefits \citep{ruxton2004obligate}. Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of around 80 kg.

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge. Their visual acuity is famously poor and echolocation does not lend itself to discovering immobile carrion. Their small size and poor terrestrial ability would also count against them at a carcass. The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle.

\section*{Terrestrial Scavengers}
A simplificaiton of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly. They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence, they have a much more restricted view of the landscape than do aerial foragers. No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent. \cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so. The same authors calculated that "a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

@@

Similar to vultures they have well developed sensory organs, particularly in olfaction whereby they can detect a rotting carcass 2 km downwind. They have a characteristic "rocking horse gait"  which allows them to cover great distances efficiently. The bone crushing ability of hyenas reveals another useful scavenger trait. Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over. Being able to extract nutrients from remnants gives the scavenger a great advantage. Osteophagy is known across a range of terrestrial carnivores. Some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, making skeletal remains an enticing resource \citep{brown1989study}. This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}. Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology. As a result the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}. A major difficulty for terrestrial scavengers is competition with vultures. Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.

The order Carnivora sees its origins in the Eocene where it split into the Caniforma and Feliforma. And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores. Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past. Indeed, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas will leave far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. The intelligence, resultant tool-use and cooperative nature of hominins means they could likely adapt to take on more or less carrion depending on their environment.

@@

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs. These animals ranged from the chicken-sized to the whale-sized all of which were bipedal. They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}. Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception. We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{depalma2013physical}. Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevelance of scavenging across a range of body sizes. In it the authors demonstrated that species of intermediate body masses (approx. 500 kg) would have gained the most benefit from scavenging. This was the result of gut capacity limitations and the effects of competition at the carcass. At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting. For the smaller species, larger competitors would have prevented their access to carrion.

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger. 
In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T rex}. There is direct evidence that \textit{T.rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}. The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}. Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}. When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it  \citep{chure1997one}.

@@

\section*{Aquatic Scavengers} 
An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In this latter low-light environment detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

The presence of occasional bounties of carrion in the form of whale falls has led some researchers to investigate if a scavenger could survive by seeking out these remains exclusively. \cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely. Any animal that could seek out such whale carcasses is unlikely to have ignored other types of carrion. Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of whales, including \textit{Leedsichthys}, a bony fish from the Jurassic Period, that weighed in excess of 20 tonnes. Thus, the energetic feasiblity of a marine scavenger has a long history. One point of interest is that of the whaling industry, which provided a bonanza of floating carcasses especially during the 20th century \citep{Whitehead415}. This meant Killer Whales \textit{Orcinus orca} could switch from hunting to scavenging, a switch made that much easier by the noise of the whaling vessels that would effectively ring the "dinner-bells" \citep{Whitehead415}. Early whales such as \textit{Basilosaurus} seem to fit into the same niche as Killer Whales and we have some evidence for scavenging in this group as well \citep{fahlke2012bite}.

@@

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{ecography}
\usepackage{lmodern}
\usepackage{amsmath}

@@

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today. Given their unique position,%TG: what do you mean by their position? Phylogenetic? Position on the "scavenging scale"?
 they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. As such, we can begin by exploring the adaptations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Moreover, having a panoramic view means being able to gather a wealth of information from other foragers, be they conspecifics or other species \citep{jackson2008effect}. Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}. These behaviours allow them forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \citep{KaneVul}.

@@

Avian flight originates in the Late Jurassic (163.5-145 Million years ago - Mya), conincident with the fossils of \textit{Archaeopteryx lithographica} so many of these benefits would have been realised from that point on for carnivorous birds. However, vertebrate flight is much older than this where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya). Scavenging in this diverse group has been hypothesied many times before. Certain clades of these animals could reach enormous sizes (e.g. Azhdarchids with wingspans of 11 metres)%CITE
 and look to have engaged in soaring flight \citep{witton2010size}. Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against their existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficency indicates they would have been comfortable foraging on the ground. Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) %TG: Check if that's the one
have a comparable morphology and are noted facultative scavengers so it is reasonable to believe that certain pterosaurs behaved similarly.

Large body size confers substantial dominance benefits \citep{ruxton2004obligate}. Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of around 80 kg. %CITE

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge. Their visual acuity is famously poor and echolocation does not lend itself to discovering immobile carrion. Their small size and poor terrestrial ability would also count against them at a carcass. The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle.
%TG: Two points here: one speculation for the non-scavenging behaviour could be because of diet. Big bats (which are better suited for scavenging, following previous paragraph) are essentially frugivores and therefore probably lack of adaptations for digesting the meat. Also carnivores bats are mainly the microbats that are insectivores. Maybe they scavenge on dead insects but here we're talking about scavenging on real meat right?
%TG: Second point: bats fossil record is of course not so good but it's actually way better than birds! Most of the data we have is on teeth of course but it's probably enough for having a rough body mass and diet estimate! And that's much better than birds where you're stuck with just some fragments of long bones.

\section*{Terrestrial Scavengers}
A simplification of terrestrial, vertebrate scavengers is one of them existing in a two-dimensional plane while foraging for carrion directly. They can detect carcasses at a range that is defined by the radius of their sensory organs, usually the visual and olfactory senses. As a consequence, they have a much more restricted view of the landscape than do aerial foragers. No contemporary terrestrial vertebrate exists as an obligate scavenger but most if not all are facultative scavengers to some extent. \cite{ruxton2004obligate} offer a reason for this in that the traits that allow for vultures to exist as scavengers undermined their ability to hunt but that the same forces have not prevented mammals from doing so. The same authors calculated that "a 1 tonne mammal or reptile, in an ecosystem yielding carrion at densities similar to the current Serengeti, could have met its energy requirements if it could detect carrion over a distance of the order of 400–500 m."\citep{ruxton2004obligate}.

@@

Similar to vultures they have well developed sensory organs, particularly in olfaction whereby they can detect a rotting carcass 2 km downwind. They have a characteristic "rocking horse gait" which allows them to cover great distances efficiently. The bone crushing ability of hyenas reveals another useful scavenger trait. Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over. Being able to extract nutrients from remnants gives the scavenger a great advantage. Osteophagy is known across a range of terrestrial carnivores. Some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, making skeletal remains an enticing resource \citep{brown1989study}. This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}. Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology. As a result the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P. brevirostris} \citep{palmqvist2011giant}. A major difficulty for terrestrial scavengers is competition with vultures. Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.

The order Carnivora sees its origins in the Middle Eocene (56-33.9 Mya) %TG: the middle Eocene date is actually for stem carnivora. It's rather important for paleao people but I think it will sound pedantic in this review
where it split into the Caniforma and Feliforma. And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores. Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past. Indeed, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae; 33.9-23.03 Mya) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene (3.6-0.126 Mya) has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas will %TG: or "could"? This is hypothetical behaviour stuff! Maybe past hyenas where exclusive sushi eaters!
 leave far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. The intelligence,%TG: Not sure about the intelligence thing: the sentence as the same meaning without and will be less anthropocentric sounding.
 resultant tool-use and cooperative nature of hominins means they could likely adapt to take on more or less carrion depending on their environment.

@@

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs. These animals ranged from the chicken-sized to the whale-sized all of which were bipedal. They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}. Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception. We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{depalma2013physical}. Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevalence of scavenging across a range of body sizes. In it the authors demonstrated that species of intermediate body masses (approx. 500 kg) would have gained the most benefit from scavenging. This was the result of gut capacity limitations and the effects of competition at the carcass. At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting. For the smaller species, larger competitors would have prevented their access to carrion.

As we discussed for the case of Cenozoic carnivores, osteophagy could be extremely beneficial to a scavenger. 
In Mesozoic systems some extremely large theropod dinosaurs had a morphology which suggests an ability to process bone e.g. the robust skull and dentition of \textit{T. rex}. There is direct evidence that \textit{T. rex} did this in the form of distinctive wear marks on its tooth apices \citep{farlow1994wear,schubert2005wear} and the presence of bone fragments in its coprolites \citep{chin1998king}. The animal also had an enormous bite force, with one estimate putting it at 57000 Newtons \citep{bates2012estimating}. This is noted as being powerful enough to break open skeletal material \citep{rayfield2001cranial}. Osteophagy may have been even more viable during this era because the body mass distribution of herbviores tended to be skewed towards larger sizes \citep{10.1371/journal.pone.0051925}. When we couple this with the fact that skeletal mass scales greater than linearly with body mass \citep{prange1979scaling} there would have been a lot of bone material to consume in the environment provided an animal had the biology to process it \citep{chure1997one}.

@@

\section*{Aquatic Scavengers} 
An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial environments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4 million years old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In this latter low-light environment detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

The presence of occasional bounties of carrion in the form of whale falls has led some researchers to investigate if a scavenger could survive by seeking out these remains exclusively. \cite{ruxton2005searching} argued that although this is energetically feasible it's ecologically unlikely. Any animal that could seek out such whale carcasses is unlikely to have ignored other types of carrion. Although no aquatic species have ever exceeded the size of whales, some enormous animals have evolved in this environment before the evolution of whales, including \textit{Leedsichthys}, a bony fish from the Middle Jurassic (174.1-163.5 Mya), that weighed in excess of 20 tonnes. Thus, the energetic feasiblity of a marine scavenger has a long history. One point of interest is that of the whaling industry, which provided a bonanza of floating carcasses especially during the 20th century \citep{Whitehead415}. This meant Killer Whales \textit{Orcinus orca} could switch from hunting to scavenging, a switch made that much easier by the noise of the whaling vessels that would effectively ring the "dinner-bells" \citep{Whitehead415}. Early whales such as \textit{Basilosaurus} seem to fit into the same niche as Killer Whales and we have some evidence for scavenging in this group as well \citep{fahlke2012bite}.

@@

% %TG: LaTeX format template for the equations in the R file
% \begin{equation}
%   \text{Travelling speed} = \frac{duration \times speed}{2} \times 10^3
% \end{equation}
% where \textit{C} is a constant of $1.15$ for mammals and $0.23$ for squamates, %TG: if you meant squamates from reptiles
% and speed being:
% \begin{equation}
%   speed = C \times M^{0.12}
% \end{equation}
